---
title: "Covid91 vaccine study Final2024"
author: "Tiffany Alvarado!"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

$H_0$ The drug does effect men differently. 
$H_A$ The drug does not effect men differently. 


```{r}
barchartGC(~treatment + sex,data=FinalDat)
barchartGC(~treatment + sex,data=FinalDat, type="percent")

```
```{r}
table1 <- xtabs(~treatment + sex, data=FinalDat)
rowPerc(table1)
colPerc(table1)
```

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```
```{r}
xtabs(~sex + treatment, data =FinalData)
```

The inferential results show that the effectiveness of the treatment is nearly identical to the females but not exact with the men being higher at 9283. we would reject the null hypothesis and the drug does affect men slightly differently as well as the p-value being 0.6536 there for these males should still take the treatment.


## Females

$H_O$ The drug effects females differently. 
$H_A$ The drug does not effect females differently. 

```{r}
xtabs(~sex + treatment, data =FinalData)
```

```{r}
# Gender data
gender <- c("Guys", "Gals")
effectiveness <- c(87, 92)

# Create barplot
barplot(effectiveness, names.arg = gender, col = "blue",
        xlab = "Gender", ylab = "Effectiveness (%)",
        main = "Vaccine Effectiveness by Gender")

```

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```


the results of the chi test show that it is equal to 0.1993 and the p value is 0.8355 and there is not much of a difference between the effect on men vs women and the effects on the drug itself so as we can see from this test and the last test there is not much of a difference but still a difference so we would reject the null hypothesis

## LGBTQ

$H_0$ Vaccine effectiveness is independent of sexual orientation 
$H_A$ Vaccine effectiveness is dependent of sexual orientation

 
```{r}
xtabs(~LGBTQ + treatment, data = FinalData)
```

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

```{r}
table1 <- xtabs(~LGBTQ + treatment, data=FinalData)
rowPerc(table1)
colPerc(table1)
```
```{r}
# Sexual orientation data
orientation <- c("Heterosexual", "Non-heterosexual")
effectiveness <- c(88, 89)

# Create barplot
barplot(effectiveness, names.arg = orientation, col = "purple",
        xlab = "Sexual Orientation", ylab = "Effectiveness (%)",
        main = "Vaccine Effectiveness by Sexual Orientation")

```



What you can see between these tests are the slight visual difference of the effectiveness between gays and straights since there is almost no difference and the p-value is 0.6 we fail to reject the null hypthesis. 

## Druggies

$H_0$ Vaccine effectiveness is independent of drug use.  
$H_A$ Vaccine effectiveness is dependent on drug use.

```{r}
table1 <- xtabs(~DrugUser + treatment, data=FinalData)
rowPerc(table1)
colPerc(table1)
```

```{r}
# Drug use data
drug_use <- c("Users", "Non-users")
effectiveness <- c(70, 90)

# Create barplot
barplot(effectiveness, names.arg = drug_use, col = "green",
        xlab = "Drug Use", ylab = "Effectiveness (%)",
        main = "Vaccine Effectiveness by Drug Use")

```

```{r}
xtabs(~ DrugUser + treatment, data = FinalData)
```

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

we can see from the charts that the effectiveness for none users is higher then the effectiveness for none users and the p-value being 0.8 we fail to reject the null hypothesis. 

# Overall Results and Conclusions

all my conclusions are underneath each subheading i was very confused on this but hopefully i did it correct!!
